---
title: "Project heatmaps"
author: "Amy Ó Brolcháin"
date: '2022-07-09'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This markdown file shows how use the output of ABRicate or AMRFinder to create presence/absence matrices and how to use those matrices in order to create heatmaps using the pheatmap library.

## Setup

First, the libraries to be loaded and the data sets to be used must be defined:

```{r libraries, message=FALSE}
#dplyr is used for manipulating data
library(dplyr)
#RColourBrewer is used for its colours
library(RColorBrewer)
#pheatmap is used for the pheatmap() function to make heatmaps
library(pheatmap)

#Setting names of ABRicate and AMRFinder reports
card_data = "card_report.txt"
vfdb_data = "vfdb_report.txt"
ecoli_vf_data = "ecoli_vf_report.txt"
plasmid_data = "plasmidfinder_report.txt"
amrfinder_data = "amrfinder_report.txt"

#Setting metadata files
##Contains 3 columns: an accession column that corresponds to an identifier in the first
##coulumn of the report, a species column with the species name, and a strain coulmn with
##the strain name. See example included.
name_file = "names_total.csv"

##Heatmaps metadata file has the names used from the name file (columns 2-3) as the entries
##of the first column, and variables to examine as the next columns. See example included.
metadata = read.csv("heatmaps_metadata.csv")
rownames(metadata)=metadata$Names
metadata_heat=select(metadata, !Names)

#Setting up colours for annotation manually using hex colour codes
##These can be altered to suit the data
metadata_heat$Source = as.factor(metadata_heat$Source)
metadata_heat$Location = as.factor(metadata_heat$Location)
metadata_heat$Assembly.Status = as.factor(metadata_heat$Assembly.Status)
ann_colours = list(
  Source = c(Bird = "#f8f132", Deer_meat = "#ff9be7", Domesticated_Animal = "#a840ac", 
             Environment = "#68d165", Freshwater = "#4941ea", Human = "#e42229",
             Wildlife = "#ff8736", Forest_soil = "#58f1a5"),
  Location = c(Australia = "#f8f132", China = "#ff8736", Ireland = "#0b9013",
               Japan = "#a840ac", Pakistan = "#ff91a2", UK = "#4941ea",
               USA = "#e42229", Eastern_Europe = "#61e3ff", Western_Europe = "#c38161",
               Southern_Europe = "#cfc1f7", Northern_Europe = "#58f1a5"),
  Assembly.Status = c(Contig = "#e42229", Scaffold = "#ff8736", Complete = "#00bb00")
)
```


## Function

The next step is to set up the function that creates the copy number matrix which forms the input for the heatmaps

```{r function}
#This function creates a data frame of gene copy numbers bacteria (as columns) and genes (as rows), where each entry corresponds to the number of copies of that gene (row) found in that bacterium (column)
#The function is designed to work with ABRicate and AMRFinder datasets.
Copy_number_matrix = function(file, name_file, dis_col, dis_val, amr){
  
  #Loading the data
  foi=read.delim(file=file, header = T, sep = "\t")
  
  #Filter out objects in the file that are not of interest
  #Designed to filter out any headers
  #For ABRicate data it can be used to filter for any variable
  if(amr==F){
    interest = filter(foi,  foi[,dis_col] == dis_val)
  }
  if(amr==T){
    interest = filter(foi,  foi[,dis_col] == "+" | foi[,dis_col] == "-")
  }
  
  #Add a row with desired strain names to the data
  names=read.csv(name_file, header = T)[,1]
  strain_list=c()
  for(row in 1:nrow(interest)){
    for(mic in names){
      if(grepl(mic, interest[row,1])){
        strain_list[row]=mic
      }
    }
  }
  mutate(interest,STRAIN=strain_list)->interest
  
  #Setting up a dataframe, whith rows corresponding to unique genes, columns to bacteria
  if(amr==F){
    genes=data.frame(matrix(nrow = length(unique(interest$GENE)), ncol = length(names)))
    colnames(genes)=names
    rownames(genes)=unique(interest$GENE)
    genes[is.na(genes)]=0
  }
  if(amr==T){
    genes=data.frame(matrix(nrow = length(unique(interest[,7])), ncol = length(names)))
    colnames(genes)=names
    rownames(genes)=unique(interest[,7])
    genes[is.na(genes)]=0
  }
  
  #Populating rows of the dataframe with the number of copies of each gene
  if(amr==F){
    for(row in 1:nrow(interest)){
      rel_row = interest[row,]
      genes[rel_row$GENE,rel_row$STRAIN] = genes[rel_row$GENE,rel_row$STRAIN]+1
    }
  }
  if(amr==T){
    for(row in 1:nrow(interest)){
      rel_row = interest[row,]
      genes[rel_row[,7],rel_row$STRAIN] = genes[rel_row[,7],rel_row$STRAIN]+1
    }
  }
  
  #Setting up the name file to be used to rename the bacteria as desired
  #First column should correspond to the current column names of the matrix
  species_list=read.csv(name_file, header = T)
  row.names(species_list)=species_list[,1]
  species_list=select(species_list,!1)
  
  #Labelling by species
  curr_species=colnames(genes)
  new_names=c()
  for(col in 1:length(curr_species)){
    new_names[col] = paste0(species_list[curr_species[col],1], "_",
                            species_list[curr_species[col],2])
  }
  colnames(genes) = new_names
  
  #Returning the matrix
  return(genes)
  }
```


## Heatmaps

Heatmaps are created from each of the datasets.

### CARD

```{r card_map}
#Using the function to format the data
card = Copy_number_matrix(file = card_data,name_file = name_file,
                          dis_col = "DATABASE",dis_val = "card",amr = F)

#Creating the heatmap using the pretty heatmap function
hv = pheatmap(as.matrix(card), cellheight = 6, cellwidth = 7,
              color = colorRampPalette(rev(brewer.pal(n = 11, name = "Spectral")))(100),  
              fontsize = 8,
              fontsize_row = 6, fontsize_col = 6,
              #cluster_rows = F,
              cutree_cols = 10,
              cutree_rows = 12,
              border_color="white",
              main = "CARD genes",
              annotation_col = metadata_heat,
              annotation_colors = ann_colours,
              filename = "card_heatmap_1.pdf")
```

### VFDB

```{r VFDB_map}
#Using the function to format the data
vfdb = Copy_number_matrix(file = vfdb_data,name_file = name_file,
                          dis_col = "DATABASE",dis_val = "vfdb",amr = F)

#Creating the heatmap using the pretty heatmap function
##Genes present in more than 20 copies are set to 4, as astA is present in 34 copies in
##E. marmotae HT073016_2, and no other gene is present in more than 4 copies
vfdb_red = vfdb
vfdb_red[vfdb_red>33] = 4
hv = pheatmap(as.matrix(vfdb_red), cellheight = 6, cellwidth = 7,
              color = colorRampPalette(
                rev(c((brewer.pal(n = 9, name = "Spectral")),"#1E1E66")))(100),  
              fontsize = 8,
              fontsize_row = 6, fontsize_col = 6,
              cutree_cols = 8,
              cutree_rows = 26,
              border_color="white",
              main = "VFDB genes",
              annotation_col = metadata_heat,
              annotation_colors = ann_colours,
              filename = "vfdb_heatmap_1.png")
```

### E. coli VF

```{r ecolivf_map}
#Using the function to format the data
ecoli_vf = Copy_number_matrix(file = ecoli_vf_data,name_file = name_file,
                             dis_col = "DATABASE",dis_val = "ecoli_vf",amr = F)

#Creating the heatmap using the pretty heatmap function
##Genes present in more than 20 copies are set to 4, as astA is present in 34 copies in
##E. marmotae HT073016_2, and no other gene is present in more than 4 copies
ecoli_red = ecoli_vf
ecoli_red[ecoli_red>33] = 4
hv = pheatmap(as.matrix(ecoli_red), cellheight = 6, cellwidth = 7,
              color = colorRampPalette(
                rev(c((brewer.pal(n = 11, name = "Spectral")),"#1E1E66")))(100),  
              fontsize = 8,
              fontsize_row = 6, fontsize_col = 6,
              cutree_cols = 10,
              cutree_rows = 35,
              border_color="white",
              main = "E. coli VF genes",
              annotation_col = metadata_heat,
              annotation_colors = ann_colours,
              filename = "ecoli_vf_heatmap_1.png")
```

### PlasmidFinder

```{r plasmid_map}
#Using the function to format the data
plasmids = Copy_number_matrix(file = plasmid_data,name_file = name_file,
                              dis_col = "DATABASE",dis_val = "plasmidfinder",amr = F)

#Creating the heatmap using the pretty heatmap function
hv = pheatmap(as.matrix(plasmids), cellheight = 7, cellwidth = 7,
              color = colorRampPalette(rev(brewer.pal(n = 11, name = "Spectral")))(100),  
              fontsize = 8,
              fontsize_row = 6, fontsize_col = 6,
              cutree_cols = 15,
              border_color="white",
              main = "PlasmidFinder genes",
              annotation_col = metadata_heat,
              annotation_colors = ann_colours,
              filename = "plasmidfinder_heatmap_1.pdf")
```

### AMRFinder

```{r amrfinder_map}
#Using the function to format the data
amrfinder = Copy_number_matrix(file = amrfinder_data,name_file = name_file,
                              dis_col = "Strand",dis_val = NA,amr = T)

#Creating the heatmap using the pretty heatmap function
hv = pheatmap(as.matrix(amrfinder), cellheight = 7, cellwidth = 7,
              color = colorRampPalette(
                rev(c(brewer.pal(n = 8, name = "Spectral"),"#1E1E66")))(100),  
              fontsize = 8,
              fontsize_row = 6, fontsize_col = 6,
              cutree_cols = 12,
              cutree_rows = 25,
              border_color="white",
              main = "AMRFinder genes",
              annotation_col = metadata_heat,
              annotation_colors = ann_colours,
              filename = "amrfinder_heatmap_1.pdf")
```

